Lonely and online: an investigation of social media use and social isolation in China

### 2025-03-17 firstmonday/Social Isolation in China fm.qmd
Reviewer responses

Most of the existing research on the subject focuses on Western highly developed societies, and this focus limits variation on key independent variables given the similar social and institutional context in which users are embedded. To address this problem, this study aims to understand how these complex dynamics interact in the non-Western setting of China. We believe the setting of China gives additional leverage to investigate issues of heterogeneity and the importance of context for several reasons. Firstly, the highly controlled nature of China's internet space prevents user exposure to some of the more threatening online content (such as cyberbullying) that is hypothesized to be a factor in creating a negative relationship between social media and feelings of social isolation [@wang2020]. If this mechanism were the dominant reason for previous findings of a negative relationship between the two, one would expect the relationship to be more positive in China. Relatedly, China has a more commercial and robust influencer culture that often dominates user's social media consumption [@tam2019; @wei2023]. If the main pathway creating a negative relationship is via user's comparisons of their lives with others (FOMO), we expect the relationship between social media use and feelings of social isolation to be greater in China.

@@

Finally, our results also further suggest more work needs to be done in identifying important heterogeneous effects - in our results, those living in a big city experienced almost no change in social isolation as social media use increased while the relationship becomes increasingly negative by ruralness. This specific heterogeneous effect may be unique to China but the strength of it suggests that important inter-personal factors may be relevant in other contexts. Importantly, given our survey results are from 2015, our finding of important heterogeneous effects suggest that these effects did not begin around the time of recent scholarly interest in the topic [@beyens2020; @ostic2021] but rather have been an important feature of the relationship between social media and social isolation since the beginning of social media use.

>>>>

Most of the existing research on the subject focuses on Western highly developed societies, and this focus limits variation on key independent variables given the similar social and institutional context in which users are embedded. To address this problem, this study aims to understand how these complex dynamics interact in the non-Western setting of China. We believe the setting of China gives additional leverage to investigate issues of heterogeneity and the importance of context for several reasons. Firstly, China's internet is highly controlled, not only by blocking foreign websites, but also by much more hands-on regulation of what the Chinese government deems unhealthy content [@wang2020]. While cyberbullying and antisocial behavior is a regular phenomenon on the Chines Internet, the government has been more proactive in trying to regulate it than in the West [@huang2021; @surveyd2023]. One expectation is therefore that Chinese social media space generates fewer feelings of social isolation - though, if the government's efforts have been not been successful, it is plausible that there is no difference compared to studies in the West. On the other hand, an effect that generates a prediction in the opposite direction is that China has a more commercial and robust influencer culture that often dominates user's social media consumption [@tam2019; @wei2023]. If the main pathway creating a negative relationship is via user's comparisons of their lives with others (FOMO), we expect the relationship between social media use and feelings of social isolation to be greater in China.

@@

One finding that matches the cyberoptimist theory of social media is that users who use social media to make new connections do report less social isolation. As @clark2018 found, the motivation for social media use matters. In their study, users who go onto social media sites for activities such as "social snacking" or quick but shallow satisfaction of social needs report higher degrees of social isolation. This is opposed to users who explicitly use social media to build communities of like-minded individuals. While our dataset does not provide information on intentionality, our findings here are in line with this kind of differential effect.

The interaction effects in our model also further suggests more work needs to be done in identifying important heterogeneous effects - in our results, those living in a big city experienced almost no change in social isolation as social media use increased while the relationship becomes increasingly negative by ruralness. This specific heterogeneous effect may be unique to China but the strength of it suggests that important inter-personal factors may be relevant in other contexts. Importantly, given our survey results are from 2015, our finding of important heterogeneous effects suggest that these effects did not begin around the time of recent scholarly interest in the topic [@beyens2020; @ostic2021] but rather have been an important feature of the relationship between social media and social isolation since the beginning of social media use.
